Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-01-13 paper.qmd
Try to find a better model for Hand-written digit with different number of bins and new choice  with tSNE perplexity 89

nldr_mnist1 <- tsne_mnist |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1, colour='#999999') +
  interior_annotation("a", c(0.08, 0.93)) +
  theme(aspect.ratio = 1)

@@

nldr_mnist6 <- tsne_mnist2 |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1, colour='#984ea3') +
  interior_annotation("f", c(0.08, 0.93)) +
  theme(aspect.ratio = 1)

@@

error_plot_mnist <- ggplot(error_mnist, 
                          aes(x = a1, 
                              y = MSE, 
                              colour = method)) + 
  geom_point(size = 0.8) +
  geom_line(linewidth = 0.3) + 
  # geom_vline(xintercept = 15, linetype="solid", 
  #            color = "black", linewidth=0.8, alpha = 0.5) +
  scale_x_continuous(breaks = sort(unique(error_mnist$a1))[append(seq(1, length(unique(error_mnist$a1)), by = 5), 24)]) +
  scale_color_manual(values=c('#e41a1c','#377eb8',"#ff7f00", '#999999', '#984ea3', '#a65628')) +
  scale_y_log10() +
  ylab("log(MSE)") +
  xlab(expression(paste("binwidth (", a[1], ")"))) +
  theme_minimal() +
  theme(panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 12, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        legend.position = "none",
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7))

```{r}
#| fig-cap: "Assessing which of the 5 NLDR layouts on the MNIST digit 1 data is the better representation using MSE for varying binwidth ($a_1$). Colour used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-e). All the layouts appear to be very similar. Layout c is universally poor. Layouts a that show two close clusters are universally suboptimal. Layout a is the best choice."
#| label: fig-mnist-mse
#| fig-pos: H
#| out-width: 80%

@@

mnist_scaled_obj <- gen_scaled_data(
  data = tsne_mnist)
tsne_minst_scaled <- mnist_scaled_obj$scaled_nldr

@@

<!-- Fit the model and compute error-->
```{r}
#| label: hexbin-mnist
## Compute hexbin parameters
num_bins_x_mnist <- 19
lim1 <- mnist_scaled_obj$lim1
lim2 <- mnist_scaled_obj$lim2
r2_mnist <- diff(lim2)/diff(lim1)

mnist_model <- fit_highd_model(
  training_data = training_data_mnist,
  emb_df = tsne_minst_scaled,
  bin1 = num_bins_x_mnist,
  r2 = r2_mnist,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "PC"
)

@@

benchmark_mnist <- 0.1

@@

error_df <- error_df |>
  mutate(sqrt_row_wise_abs_error = sqrt(row_wise_abs_error))

highd_error_2d_mnist <- error_df |>
  ggplot(aes(x = tSNE1,
             y = tSNE2,
             colour = sqrt_row_wise_abs_error)) +
  geom_point(alpha=0.2) +
  scale_colour_continuous_sequential(palette = "YlOrRd", n_interp = 20) +
  geom_rect(aes(xmin = 0.8, xmax = 0.9, ymin = 0.45, ymax = 0.6),
               fill = "transparent", color = "#000000", linewidth = 0.7) +
  geom_rect(aes(xmin = 0.75, xmax = 0.85, ymin = 0.8, ymax = 0.95),
               fill = "transparent", color = "#e41a1c", linewidth = 0.7)  +
  geom_rect(aes(xmin = 0.45, xmax = 0.55, ymin = 0.45, ymax = 0.6),
               fill = "transparent", color = "#4daf4a", linewidth = 0.7)  +
  geom_rect(aes(xmin = 0.7, xmax = 0.8, ymin = 0.25, ymax = 0.4),
               fill = "transparent", color = "#984ea3", linewidth = 0.7) 
  # scale_colour_brewer(
  #   type = "seq",
  #   palette = 1
  # ) #+ 
  # theme(
  #     legend.position = "bottom",
  #     legend.text = element_text(size = 10),          # Increase legend text size
  #     legend.key.size = unit(1, "lines")              # Increase size of legend keys
  # ) +
  # guides(
  #     color = guide_legend(
  #         override.aes = list(size = 2.5, alpha = 1)      # Increase point size and opacity in legend
  #     )
  # ) 
```

@@

mnist_proj_tsne_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2,
      colour = cluster)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000") +
  geom_point(
    #size = 0.2,
    alpha = 0.3) +
  scale_color_manual(values = c("#999999",'#ff7f00')) +
  geom_segment(
    data=axes,
    aes(x=x1, y=y1, xend=x2, yend=y2),
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2), 
    label=rownames(axes),
    colour="grey50",
    size = 2) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  #coord_fixed() +
  xlim(c(-1, 0.6)) +
  ylim(c(-1, 0.6)) +
  interior_annotation("b") +
  theme(
    legend.position = "none"
  )

>>>>

nldr_mnist1 <- tsne_mnist |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1, colour='#984ea3') +
  interior_annotation("a", c(0.08, 0.93)) +
  theme(aspect.ratio = 1)

@@

nldr_mnist6 <- tsne_mnist2 |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1, colour='#999999') +
  interior_annotation("f", c(0.08, 0.93)) +
  theme(aspect.ratio = 1)

@@

error_plot_mnist <- ggplot(error_mnist, 
                          aes(x = a1, 
                              y = MSE, 
                              colour = method)) + 
  geom_point(size = 0.8) +
  geom_line(linewidth = 0.3) + 
  # geom_vline(xintercept = 15, linetype="solid", 
  #            color = "black", linewidth=0.8, alpha = 0.5) +
  scale_x_continuous(breaks = sort(unique(error_mnist$a1))[append(seq(1, length(unique(error_mnist$a1)), by = 5), 24)]) +
  scale_color_manual(values=c('#e41a1c','#377eb8',"#ff7f00", '#984ea3', '#999999', '#a65628')) +
  scale_y_log10() +
  ylab("log(MSE)") +
  xlab(expression(paste("binwidth (", a[1], ")"))) +
  theme_minimal() +
  theme(panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 12, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        legend.position = "none",
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7))

```{r}
#| fig-cap: "Assessing which of the 5 NLDR layouts on the MNIST digit 1 data is the better representation using MSE for varying binwidth ($a_1$). Colour used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-e). All the layouts appear to be very similar. Layout c is universally poor. Layouts f that show two close clusters are universally suboptimal. Layout f is the best choice."
#| label: fig-mnist-mse
#| fig-pos: H
#| out-width: 80%

@@

mnist_scaled_obj <- gen_scaled_data(
  data = tsne_mnist2)
tsne_minst_scaled <- mnist_scaled_obj$scaled_nldr

@@

<!-- Fit the model and compute error-->
```{r}
#| label: hexbin-mnist
## Compute hexbin parameters
num_bins_x_mnist <- 30
lim1 <- mnist_scaled_obj$lim1
lim2 <- mnist_scaled_obj$lim2
r2_mnist <- diff(lim2)/diff(lim1)

mnist_model <- fit_highd_model(
  training_data = training_data_mnist,
  emb_df = tsne_minst_scaled,
  bin1 = num_bins_x_mnist,
  r2 = r2_mnist,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = TRUE,
  col_start_highd = "PC"
)

@@

#benchmark_mnist <- 0.1

@@

error_df <- error_df |>
  mutate(sqrt_row_wise_tot_error = sqrt(row_wise_total_error))

highd_error_2d_mnist <- error_df |>
  ggplot(aes(x = tSNE1,
             y = tSNE2,
             colour = sqrt_row_wise_tot_error)) +
  geom_point(alpha=0.2) +
  scale_colour_continuous_sequential(palette = "YlOrRd", n_interp = 20) +
  geom_rect(aes(xmin = 0.8, xmax = 0.9, ymin = 0.45, ymax = 0.6),
               fill = "transparent", color = "#000000", linewidth = 0.7) +
  geom_rect(aes(xmin = 0.75, xmax = 0.85, ymin = 0.8, ymax = 0.95),
               fill = "transparent", color = "#e41a1c", linewidth = 0.7)  +
  geom_rect(aes(xmin = 0.45, xmax = 0.55, ymin = 0.45, ymax = 0.6),
               fill = "transparent", color = "#4daf4a", linewidth = 0.7)  +
  geom_rect(aes(xmin = 0.7, xmax = 0.8, ymin = 0.25, ymax = 0.4),
               fill = "transparent", color = "#984ea3", linewidth = 0.7) 
  # scale_colour_brewer(
  #   type = "seq",
  #   palette = 1
  # ) #+ 
  # theme(
  #     legend.position = "bottom",
  #     legend.text = element_text(size = 10),          # Increase legend text size
  #     legend.key.size = unit(1, "lines")              # Increase size of legend keys
  # ) +
  # guides(
  #     color = guide_legend(
  #         override.aes = list(size = 2.5, alpha = 1)      # Increase point size and opacity in legend
  #     )
  # ) 
```

@@

mnist_proj_tsne_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2,
      colour = cluster)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000") +
  geom_point(
    #size = 0.2,
    alpha = 0.1) +
  scale_color_manual(values = c("#999999",'#ff7f00')) +
  geom_segment(
    data=axes,
    aes(x=x1, y=y1, xend=x2, yend=y2),
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2), 
    label=rownames(axes),
    colour="grey50",
    size = 2) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  #coord_fixed() +
  xlim(c(-1, 0.6)) +
  ylim(c(-1, 0.6)) +
  interior_annotation("b") +
  theme(
    legend.position = "none"
  )
